---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
mydata <- read.csv("activity.csv")
mydata$date <- as.Date(mydata$date)
cleanNA <- subset(mydata, !is.na(mydata$steps))
```


## Histogram of the total number of steps each day
```{r sumbyday}
library(ggplot2)
totalEachDay <- tapply(cleanNA$steps, cleanNA$date, sum)
qplot(totalEachDay, ylab = "Frequency", xlab = "Total steps per day", binwidth = 500)
```

## What is mean total number of steps taken per day?

```{r meanacrossdays}
meanAcrossDays <- mean(totalEachDay)
```
The mean number of steps taken per day: `r meanAcrossDays`


## The median total number of steps taken per day:
```{r medianacrossdays}
medianAcrossDays <- median(totalEachDay)
```
The median number of steps taken per day: `r medianAcrossDays`

## What is the average daily activity pattern?

```{r}
avgInterval <-tapply(cleanNA$steps, cleanNA$interval, mean, simplify = TRUE)
df_new <- data.frame(avg = avgInterval, interval = as.integer(names(avgInterval)))
plot(df_new$interval, df_new$avg, type = "l", xlab = "5-minute interval" , ylab = "Average Number Of Steps Across Days")

```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
